---
title: "Correlation analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Correlation analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(bqreport)
```

```{r}
data <- tibble::tibble(
  x = c(1, 2, 3, 4, 5, 6, 7, 8),
  y = c(1.2, 1.7, 3.4, 3.8, 5.3, 5.7, 7.4, 7.8),
  age = c(40, 52, 47, 60, 44, 56, 49, 63),
  arm = factor(rep(c("Control", "Treatment"), each = 4))
) |>
  as_bq_data() |>
  set_role(arm, "stratum")

result <- data |>
  plan_correlations(x, with = y, adjust_for = age,
    method = pearson_correlation()) |>
  validate_plan(data) |>
  run_analysis(data)

correlations(result)
table_body(tbl_correlation(result))
```

Built-in methods include Pearson, Spearman, Kendall, biweight, weighted
Pearson, repeated-measures, polychoric, and tetrachoric correlations.

## Resampling inference

```{r}
resampled <- resampled_correlation(
  pearson_correlation(), bootstrap = 49, permutations = 49, seed = 2026
)

resampled_result <- data |>
  plan_correlations(x, with = y, method = resampled) |>
  validate_plan(data) |>
  run_analysis(data)

correlations(resampled_result)[, c("estimate", "std_error", "conf_low",
  "conf_high", "p_value", "bootstrap_successful", "permutation_successful")]
```

The seed and successful replicate counts are recorded without changing the
global RNG state.

## Strata and interaction

```{r}
stratified <- data |>
  plan_correlations(x, with = y, strata = arm,
    interaction_test = TRUE, adjust = "holm") |>
  validate_plan(data) |>
  run_analysis(data)

correlations(stratified)
tests(stratified)
contrasts(stratified)
```

The omnibus test and pairwise differences compare correlations across strata;
they are distinct from merely reporting separate stratified estimates.
