---
title: "Custom functions, methods, and selectors"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Custom functions, methods, and selectors}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(bqreport)
```

Extensions are introduced through constructors with validated outputs. The
package does not infer the meaning of an arbitrary return value.

## A custom correlation method

```{r}
cosine_correlation <- correlation_method(
  id = "cosine_similarity",
  effect_measure = "cosine_similarity",
  scale = "minus_one_to_one",
  ci_method = "none",
  supports_strata = TRUE,
  compute = function(context) {
    estimate <- sum(context$x * context$y) /
      sqrt(sum(context$x^2) * sum(context$y^2))
    correlation_output(
      estimate = estimate,
      std_error = NA_real_, std_error_scale = "similarity",
      conf_low = NA_real_, conf_high = NA_real_,
      statistic = NA_real_, df = NA_real_, p_value = NA_real_
    )
  }
)

data <- as_bq_data(tibble::tibble(x = 1:8, y = c(1, 3, 2, 5, 4, 7, 6, 8)))
result <- data |>
  plan_correlations(x, with = y, method = cosine_correlation) |>
  validate_plan(data) |>
  run_analysis(data)
correlations(result)
```

The callback receives only the prepared analysis context. Its identity and
hash are written to the plan and provenance.

## A complete custom analysis function

This example estimates relative risks with a Poisson log-link model and robust
interpretation declared in advance.

```{r}
custom_rr <- analysis_function(
  id = "custom_relative_risk",
  effect_measure = "risk_ratio",
  scale = "ratio",
  run = function(context) {
    fit <- stats::glm(context$formula, data = context$model_frame,
      family = stats::poisson("log"))
    sm <- summary(fit)$coefficients
    critical <- stats::qnorm((1 + context$confidence_level) / 2)
    analysis_output(
      model = fit,
      estimates = tibble::tibble(
        analysis_id = context$analysis_id,
        outcome = context$outcome_spec$name,
        predictor = context$predictor_spec$name,
        term = rownames(sm), level = NA_character_,
        estimate = exp(unname(sm[, "Estimate"])),
        std_error = unname(sm[, "Std. Error"]),
        std_error_scale = "log_risk",
        conf_low = exp(unname(sm[, "Estimate"] - critical * sm[, "Std. Error"])),
        conf_high = exp(unname(sm[, "Estimate"] + critical * sm[, "Std. Error"])),
        statistic = unname(sm[, "z value"]), df = NA_real_,
        p_value = unname(sm[, "Pr(>|z|)"]),
        effect_measure = "risk_ratio", scale = "ratio",
        n = as.integer(stats::nobs(fit)),
        n_events = as.integer(sum(context$response == 1)),
        method = "custom_relative_risk", variance = "model_based"
      )
    )
  }
)

binary <- tibble::tibble(
  outcome = c(0, 0, 1, 0, 1, 1, 1, 0),
  exposed = c(0, 0, 0, 1, 1, 1, 1, 0)
) |>
  as_bq_data() |>
  set_outcome(outcome, type = "binary", event = 1) |>
  set_predictor(exposed, type = "continuous")

rules <- analysis_rules(where_binary() ~ custom_rr)
custom_result <- binary |>
  plan_analysis(outcome, exposed, rules = rules) |>
  validate_plan(binary) |>
  run_analysis(binary)

estimates(custom_result)
provenance <- custom_result$provenance
provenance[, c("function_id", "function_hash")]
```

Malformed output becomes a typed issue; the runtime does not guess missing
columns or silently switch engines.

## An explicit fallback chain

A runtime fallback is available only as an inspectable chain. Each next method
must have the same estimand and output scale, and is attempted only for a
condition class listed in `advance_on`.

```{r}
primary_rr <- analysis_function(
  id = "primary_rr",
  effect_measure = "risk_ratio",
  scale = "ratio",
  run = function(context) {
    stop(structure(
      list(message = "Numerical method did not converge", call = NULL),
      class = c("bq_error_numerical_failure", "error", "condition")
    ))
  },
  fallback = list(poisson_rr = custom_rr),
  advance_on = "bq_error_numerical_failure"
)

fallback_result <- binary |>
  plan_analysis(
    outcome, exposed,
    rules = analysis_rules(where_binary() ~ primary_rr)
  ) |>
  validate_plan(binary) |>
  run_analysis(binary)

attempts(fallback_result)
fallback_result$provenance[, c(
  "method_chain", "fallback_conditions", "executed_method", "fallback_used"
)]
```

`bq_error_invalid_engine_output` and other contract violations never advance
the chain. They indicate an implementation bug rather than an expected
statistical failure.

## A custom model in a three-group workflow

The next model keeps the ordinary least-squares estimand but replaces the
model-based covariance matrix with HC3 robust standard errors. Because the
method returns the stable estimates schema, it can be used by the same planner
and reporting layer as built-in models.

```{r}
hc3_linear_model <- analysis_method(
  id = "hc3_linear_model",
  effect_measure = "mean_difference",
  scale = "identity",
  required_packages = "sandwich",

  fit = function(context) {
    stats::lm(
      context$formula,
      data = context$model_frame,
      na.action = stats::na.omit
    )
  },

  tidy_estimates = function(fit, context) {
    beta <- stats::coef(fit)
    covariance <- sandwich::vcovHC(fit, type = "HC3")
    standard_error <- sqrt(diag(covariance))
    degrees <- stats::df.residual(fit)
    critical <- stats::qt(
      (1 + context$confidence_level) / 2,
      df = degrees
    )
    statistic <- beta / standard_error

    tibble::tibble(
      analysis_id = context$analysis_id,
      outcome = context$outcome_spec$name,
      predictor = context$predictor_spec$name,
      term = names(beta),
      level = NA_character_,
      estimate = unname(beta),
      std_error = unname(standard_error),
      std_error_scale = "identity",
      conf_low = unname(beta - critical * standard_error),
      conf_high = unname(beta + critical * standard_error),
      statistic = unname(statistic),
      df = rep(degrees, length(beta)),
      p_value = unname(
        2 * stats::pt(abs(statistic), df = degrees, lower.tail = FALSE)
      ),
      effect_measure = "mean_difference",
      scale = "identity",
      n = rep(as.integer(stats::nobs(fit)), length(beta)),
      n_events = NA_integer_,
      method = "hc3_linear_model",
      variance = "HC3"
    )
  },

  tidy_tests = function(fit, context) {
    omnibus <- stats::anova(fit)
    predictor_row <- 1L

    tibble::tibble(
      analysis_id = context$analysis_id,
      outcome = context$outcome_spec$name,
      predictor = context$predictor_spec$name,
      contrast = NA_character_,
      numerator = NA_character_,
      denominator = NA_character_,
      test = "predictor_omnibus",
      statistic = unname(omnibus$`F value`[[predictor_row]]),
      df = unname(omnibus$Df[[predictor_row]]),
      p_value = unname(omnibus$`Pr(>F)`[[predictor_row]]),
      p_adjusted = NA_real_,
      adjust_method = "none",
      method = "hc3_linear_model"
    )
  },

  compute_contrasts = function(fit, context) {
    group_name <- context$predictor_spec$name[[1]]
    group_levels <- levels(context$model_frame[[group_name]])
    pairs <- t(utils::combn(seq_along(group_levels), 2L))

    reference_grid <- stats::setNames(
      list(factor(group_levels, levels = group_levels)),
      group_name
    ) |>
      as.data.frame()
    design <- stats::model.matrix(
      stats::delete.response(stats::terms(fit)),
      reference_grid
    )

    beta <- stats::coef(fit)
    covariance <- sandwich::vcovHC(fit, type = "HC3")
    degrees <- stats::df.residual(fit)
    critical <- stats::qt(
      (1 + context$confidence_level) / 2,
      df = degrees
    )

    contrast_matrix <- t(vapply(
      seq_len(nrow(pairs)),
      function(i) design[pairs[i, 2L], ] - design[pairs[i, 1L], ],
      numeric(ncol(design))
    ))
    estimate <- as.numeric(contrast_matrix %*% beta)
    standard_error <- sqrt(diag(
      contrast_matrix %*% covariance %*% t(contrast_matrix)
    ))
    statistic <- estimate / standard_error
    p_value <- 2 * stats::pt(
      abs(statistic), df = degrees, lower.tail = FALSE
    )
    comparison <- context$contrast_specs[1, , drop = FALSE]
    adjust <- comparison$adjust_method[[1]]
    numerator <- group_levels[pairs[, 2L]]
    denominator <- group_levels[pairs[, 1L]]

    tibble::tibble(
      analysis_id = context$analysis_id,
      outcome = context$outcome_spec$name,
      predictor = group_name,
      contrast_id = comparison$contrast_id[[1]],
      contrast = paste(numerator, "vs", denominator),
      numerator = numerator,
      denominator = denominator,
      estimate = estimate,
      std_error = standard_error,
      std_error_scale = "identity",
      conf_low = estimate - critical * standard_error,
      conf_high = estimate + critical * standard_error,
      p_value = p_value,
      p_adjusted = stats::p.adjust(p_value, method = adjust),
      adjust_method = adjust,
      effect_measure = "mean_difference",
      scale = "identity"
    )
  }
)
```

The model is then registered through an ordinary analysis rule.

```{r}
three_groups <- tibble::tibble(
  arm = factor(
    rep(c("Control", "Treatment A", "Treatment B"), each = 8),
    levels = c("Control", "Treatment A", "Treatment B")
  ),
  score = c(
    42, 45, 48, 50, 51, 53, 54, 57,
    47, 49, 52, 55, 57, 58, 61, 64,
    51, 54, 56, 59, 62, 64, 67, 70
  )
) |>
  as_bq_data() |>
  set_outcome(score, type = "continuous") |>
  set_predictor(arm, type = "nominal", reference = "Control") |>
  set_comparisons(arm, all_pairwise(), adjust = "holm")

hc3_plan <- three_groups |>
  plan_analysis(
    outcomes = score,
    predictors = arm,
    rules = analysis_rules(where_continuous() ~ hc3_linear_model)
  ) |>
  validate_plan(three_groups)

hc3_result <- run_analysis(hc3_plan, three_groups)

estimates(hc3_result)
tests(hc3_result)
contrasts(hc3_result)
table_body(tbl_comparison(hc3_result))
```

The model callback owns fitting and normalization, including its model-based
contrast algorithm. The registered comparison specification remains the
source of the contrast id and multiplicity adjustment. The function id, hash,
and required package version are retained in the plan and provenance.

## Data-dependent selection before fitting

Selectors choose only among declared candidates during preflight.

```{r}
sample_size_policy <- method_selector(
  id = "sample_size_policy",
  candidates = list(ordinary = linear_model(), robust = linear_model()),
  select = function(context) {
    n <- nrow(context$model_frame)
    method_choice(
      method = if (n >= 30) "ordinary" else "robust",
      reason = paste("Analyzed sample size:", n),
      diagnostics = tibble::tibble(n = n)
    )
  }
)
```

The selected concrete method, reason, diagnostics, callback hash, and package
requirements are inspectable before `run_analysis()`. Unless an explicit
method chain was declared, a fit failure never
causes the selector to run again.
