---
title: "Descriptive analysis and group comparisons"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Descriptive analysis and group comparisons}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(bqreport)
```

Descriptive templates are variable metadata. They control display rows, not
the numerical computation or rounding of an `analysis_result`.

```{r}
trial <- tibble::tibble(
  arm = factor(rep(c("Control", "Treatment"), each = 8)),
  age = c(42, 51, 63, 48, 55, 60, 46, 58, 44, 49, 54, 57, 61, 52, 47, 59),
  response = factor(
    c("No", "No", "Yes", "No", "Yes", "No", "No", "Yes",
      "Yes", "Yes", "Yes", "No", "Yes", "Yes", "No", "Yes")
  )
) |>
  as_bq_data() |>
  set_role(arm, "group") |>
  set_predictor(arm, type = "binary", reference = "Control") |>
  set_outcome(age, type = "continuous") |>
  set_outcome(response, type = "binary", event = "Yes") |>
  set_descriptive_statistics(age,
    c("{mean} ({sd})", "{median} ({q1}; {q3})", "MAD: {mad}")) |>
  set_descriptive_statistics(response, "{n}/{N} ({p}%)")

plan <- trial |>
  plan_descriptives(
    variables = c(age, response), groups = arm, overall = TRUE,
    comparisons = TRUE
  ) |>
  validate_plan(trial)

result <- run_analysis(plan, trial)
descriptives(result)
table_body(tbl_descriptive(result, locale = "en"))
```

The overall column and grouped populations are generated in the same plan.
For categorical outcomes, `n`, `N`, and `p` retain distinct numerator,
denominator, and proportion semantics.

## Custom descriptive function

Custom fields are declared before compilation and returned through a validated
tidy contract.

```{r}
coefficient_of_variation <- descriptive_function(
  id = "coefficient_of_variation",
  fields = "cv",
  types = "continuous",
  compute = function(context) {
    tibble::tibble(
      statistic = "cv",
      value = stats::sd(context$values) / mean(context$values),
      statistic_method = "sd_over_mean"
    )
  }
)

custom_data <- trial |>
  set_descriptive_statistics(age, "CV: {cv}")

custom_result <- custom_data |>
  plan_descriptives(age, groups = arm,
    functions = list(coefficient_of_variation)) |>
  validate_plan(custom_data) |>
  run_analysis(custom_data)

table_body(tbl_descriptive(custom_result))
```

`shapiro_wilk()` uses the same extension contract. Its result is diagnostic and
does not silently select a downstream model.
