---
title: "Get started with bqreport"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Get started with bqreport}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(bqreport)
```

`bqreport` treats analysis as a compilation pipeline. Data and metadata are
compiled into an inspectable plan; only a validated plan can be executed.

## Register data and metadata

```{r}
trial <- tibble::tibble(
  response = c(0, 0, 1, 0, 1, 1, 1, 0, 1, 0, 1, 1),
  treatment = factor(rep(c("Control", "Treatment"), each = 6)),
  age = c(44, 57, 51, 63, 46, 55, 60, 49, 52, 67, 58, 45),
  biomarker = c(2.1, 3.4, 2.7, 4.2, 3.0, 3.8, 4.6, 3.9, 5.1, 4.8, 5.4, 4.2)
) |>
  as_bq_data() |>
  set_outcome(response, type = "binary", event = 1) |>
  set_predictor(treatment, type = "binary", reference = "Control") |>
  set_role(treatment, "group") |>
  set_predictor(c(age, biomarker), type = "continuous") |>
  set_colors(treatment, c(Control = "#4477AA", Treatment = "#CC6677"))

variables(trial)[, c("name", "role", "type", "reference", "status")]
```

Roles are additive, so `treatment` is both a grouping variable and predictor.
Stable `var_id` values allow registry references to survive renaming.

## Compile and review a plan

```{r}
plan <- trial |>
  plan_analysis(
    outcomes = response,
    predictors = treatment,
    covariates = age
  ) |>
  validate_plan(trial)

plan[, c("analysis_id", "formula", "method", "effect_measure", "scale",
         "n_analyzed", "status")]
```

At this point the method is concrete. A data-dependent selector, if used, has
already run and recorded its reason and diagnostics.

## Execute and inspect tidy components

```{r}
result <- run_analysis(plan, trial)
estimates(result)
tests(result)
diagnostics(result)
issues(result)
```

Numbers remain unrounded. Formatting is a separate operation:

```{r}
table <- tbl_regression(result, locale = "en")
table_body(table)
```

Optional renderers convert the backend-independent table to `gt` or
`flextable`, while `plot_forest(result)` returns a normal ggplot object.
