---
title: "Longitudinal and survival workflows"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Longitudinal and survival workflows}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(bqreport)
```

## Longitudinal data

Long and wide inputs compile to the same internal long analysis frame without
mutating the original `bq_data`.

```{r eval=requireNamespace("lme4", quietly=TRUE)}
long <- expand.grid(
  id = factor(seq_len(12)), visit = factor(c("V0", "V1", "V2"),
    levels = c("V0", "V1", "V2"))
)
long$arm <- factor(ifelse(as.integer(long$id) <= 6, "Control", "Treatment"))
long$score <- 10 + as.integer(long$id) / 3 +
  as.numeric(long$visit) + (long$arm == "Treatment") * as.numeric(long$visit)

data <- tibble::as_tibble(long) |>
  as_bq_data() |>
  set_predictor(arm, type = "binary", reference = "Control") |>
  set_colors(arm, c(Control = "#4477AA", Treatment = "#CC6677")) |>
  set_longitudinal_design(id, visit, group = arm, layout = "long",
    baseline = "V0", time_scale = "categorical") |>
  add_longitudinal_outcome(score_change, values = score,
    baseline = "V0", type = "continuous")

result <- data |>
  plan_longitudinal(score_change, method = lmm_model(),
    comparisons = TRUE, adjust = "holm") |>
  validate_plan(data) |>
  run_analysis(data)

estimates(result)
contrasts(result)
```

Use `gee_model()` for a population-average estimand and `lmm_model()` for a
mixed-model estimand. The choice is explicit and never treated as interchangeable.

## Kaplan--Meier, RMST, and Cox regression

```{r eval=requireNamespace("survival", quietly=TRUE)}
survival_data <- tibble::tibble(
  time = c(3, 5, 8, 10, 4, 7, 9, 12, 2, 6, 11, 13),
  event = c(1, 0, 1, 1, 1, 0, 1, 0, 1, 1, 0, 1),
  arm = factor(rep(c("Control", "Treatment"), each = 6)),
  age = c(44, 57, 51, 63, 46, 55, 60, 49, 52, 67, 58, 45)
) |>
  as_bq_data() |>
  set_predictor(arm, type = "binary", reference = "Control") |>
  set_predictor(age, type = "continuous") |>
  set_colors(arm, c(Control = "#4477AA", Treatment = "#CC6677")) |>
  add_survival_outcome(os, time, event, event_value = 1, time_unit = "months")

km <- survival_data |>
  plan_kaplan_meier(os, groups = arm, times = c(3, 6, 9),
    quantiles = c(.25, .5), rmst_tau = 9,
    estimates = c("survival", "cumulative_risk")) |>
  validate_plan(survival_data) |>
  run_analysis(survival_data)

survival_estimates(km)

cox <- survival_data |>
  plan_survival(os, arm, covariates = age) |>
  validate_plan(survival_data) |>
  run_analysis(survival_data)

estimates(cox)
diagnostics(cox)
```

Competing-risk outcomes use `add_competing_risk_outcome()` and
`plan_cumulative_incidence()`, which estimate Aalen--Johansen cumulative
incidence rather than substituting `1 - S(t)`.
