---
title: "Regression models, transformations, and contrasts"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Regression models, transformations, and contrasts}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(bqreport)
```

```{r}
trial <- tibble::tibble(
  score = c(10, 12, 13, 15, 17, 18, 20, 22, 23, 25, 27, 30),
  arm = factor(rep(c("Control", "Low", "High"), each = 4),
    levels = c("Control", "Low", "High")),
  age = c(40, 51, 47, 60, 42, 55, 49, 63, 45, 53, 58, 61)
) |>
  as_bq_data() |>
  set_outcome(score, type = "continuous") |>
  set_predictor(arm, type = "nominal", reference = "Control") |>
  set_predictor(age, type = "continuous") |>
  set_comparisons(arm, against_reference("Control"), adjust = "holm") |>
  set_colors(arm, c(Control = "#4477AA", Low = "#DDCC77", High = "#CC6677"))

plan <- trial |>
  plan_analysis(score, arm, covariates = age) |>
  validate_plan(trial)
result <- run_analysis(plan, trial)

estimates(result)
contrasts(result)
table_body(tbl_comparison(result))
```

Available target specifications include `all_pairwise()`,
`against_reference()`, `consecutive_comparisons()`, and
`against_global_mean()`. Model coding and target comparisons remain separate.

## Explicit nonlinear covariates

Transformations and model terms are metadata and are compiled into the plan.

```{r eval=FALSE}
trial <- trial |>
  set_transformation(age, log10_transform()) |>
  set_model_term(age, natural_spline(df = 3))

plan <- plan_analysis(trial, score, arm, covariates = age)
```

This represents a sequential log transformation followed by a spline basis.
The transformation identity and basis settings are included in provenance.

## Other outcome families

System defaults cover linear, logistic, Poisson, proportional-odds ordinal,
and baseline-category multinomial models. Alternatives can be selected with
rules, for example:

```{r eval=FALSE}
rules <- analysis_rules(
  where_count() ~ negative_binomial_model(exponentiate = TRUE)
)

plan_analysis(data, outcomes = events, predictors = treatment, rules = rules)
```

Exponentiation is always explicit in the selected method specification; SEs
remain on their declared model scale.
